Reproducible Research: Peer Assessment 2
==========================================
Created by Simon Mackinnon on `r format(Sys.Date(), "%B %d, %Y")`
[Reproducable Research - Coursera](https://www.coursera.org/course/repdata)

## Economic and Public Health Effects of Storms and Other Severe Weather Events 

### Synopsis

The aim of this report is to analyse the storm database collected via the U.S. National Oceanic and Atmospheric Administration's (NOAA) from 1950 - 2011, and conclude the economic and health effects of different severe weather event types.  
The recorded data lists the number of fatalities, injuries, property and crop damage. These will be used to decide which types of event are most harmful to the population health and economy.  
In analysing the data, it was found that **tornadoes have the greatest detrimental effect on population health** and that **flood and drought have the greatest detremental economic effect**.  


### Settings
```{r settings}
options(scipen = 999) 
library(R.utils)
library(lubridate)
library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)
library(lazyeval)
```

### Data Processing
The data to be used for this analysis is available for download via the course website at [Storm Data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).    

```{r downloadingAndExtracting, echo=TRUE}
zip_filename <- "./data/repdata-data-StormData.csv.bz2"
data_filename <- "./data/repdata-data-StormData.csv"

if (!file.exists("./data"))
{
  dir.create("./data")
}

if (!file.exists(data_filename))
{
  if (!file.exists(zip_filename))
  {
    url <- "http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
    #record the time/location of the downloaded file
    print (paste("Downloading file from:", url ,"at:", format(Sys.Date(), "%B %d, %Y"), format(Sys.time(), "%H:%M:%S")))
    #changed url to be http, as download file has issues with https
    
    download.file(url, 
                  destfile = zip_filename,
                  mode="wb")
  }
  
  bunzip2(zip_filename, overwrite=T, remove=F)
}

print (paste("Extracted data file information.", "Created on:", format(file.info(data_filename)$ctime, "%B %d, %Y %H:%M:%S"), "Modified on: ", format(file.info(data_filename)$mtime, "%B %d, %Y %H:%M:%S")))
```

We now read the .csv into R.

```{r dataReading, echo=TRUE}
# this may take some time!
stormData <- read.csv(data_filename, 
                      header=TRUE, 
                      sep=",",
                      stringsAsFactors=FALSE)

#confirm the number dimensions of the read data
dim(stormData)


#and display the head rows
head(stormData)

```
There are **`r nrow(stormData)` rows** and **`r ncol(stormData)` columns** in the read data.


### Data Selection and Filtering

Looking at the details of how the data was collected at the [Storm Events Database](http://www.ncdc.noaa.gov/stormevents/details.jsp), we can see that only observations from 1996 onward includes data from *All Event Types (48 from Directive 10-1605)*, (as defined in [NWS Directive 10-1605](http://www.ncdc.noaa.gov/stormevents/pd01016005curr.pdf)).  

We can verify that most of the observations were recorded after this date as well, by observing the frequency of observations per year in the **stormData$BGN_DATE** variable

```{r observationFrequencyPerYear, echo=TRUE, fig.align='center', fig.width=10, fig.height=5}

year <- year(as.Date(stormData$BGN_DATE, format = "%m/%d/%Y"))

#break up the year data into 3 types
yearType <- sapply(as.numeric(year), function(x) 
    if (x >= 1950 & x <= 1955){1} 
    else if (x <= 1996) {2} 
    else if (x <= 2011) {3}
    else {0})

#bind into a marix, then coerce to a data frame
yearData <- data.frame(cbind(year, yearType))

#create the histogram
ggplot(yearData, aes(x=year)) + 
geom_histogram(binwidth = 1, 
               aes(fill = factor(yearData$yearType)),
               colour = "black") +
scale_fill_discrete(name="Year Type", 
                    labels= c("1. Tornado\n1950-1955\n",
                              "2. Tornado,\nThunderstorm Wind\nand Hail\n1956-1995\n",
                              "3. All Event\nTypes (48 from\nDirective 10-1605)\n1996-2011")) +
labs(title = "Histogram of Number of\nObservarions Recorded Per Year (1950 - 2011)",
     x = "Year",
     y = "Number of Observations Recorded")


```

Based on this, we filter the data to only use observations from 1996 onward.

```{r dataFiltering, echo=TRUE}

#append the year values for filtering
stormData$year <- year
#subset the data for all values of year greater or equal to 1996
stormData <- stormData[stormData$year >= 1996,]
#display the dimensions of the filtered data
dim(stormData)

```
After filtering, there are **`r nrow(stormData)` rows** and **`r ncol(stormData)` columns** in the filtered data.

### Results

#### Event Type Analysis

In order to perform an anlysis on the health and economic effects of different severe weather event types, we should first illustrate a proper understanding of the different types of events.

```{r EventTypeAnalysis1, echo=TRUE}

eventTypeList <- unique(stormData$EVTYPE)
eventTypeList

```

From this, we can see there are **`r length(eventTypeList)`** different defined event types. However, we can observe some problems with the way that the event types are recorded.  
In particular, we can see several types of severe weather that is defined by multiple EVTYPE values, e.g. Flood:



```{r EventTypeAnalysis2, echo=TRUE}

floodList <- eventTypeList[grepl("flood", eventTypeList, ignore.case = TRUE)]
floodList

```

From this, we can see there are **`r length(floodList)`** different event types that could be defined more broadly as 'flooding'. Other examples of this are tornadoes, cold, hot, dry, snow, fire, and others.  

Another problem with EVTYPE data is that there are many observations with "Summary of Month Date" recorded, instead of a definition of the event type.

```{r EventTypeAnalysis3, echo=TRUE}

evtypeSummaryOfMonthDateList <- eventTypeList[grepl("summary of", eventTypeList, ignore.case = TRUE)]
evtypeSummaryOfMonthDateList

```

From this, we can see there are **`r length(evtypeSummaryOfMonthDateList)`** observations where the event type was recorded as such. Exploring a little further, we can see how the observation was recorded in full.

```{r EventTypeAnalysis4, echo=TRUE}

head(stormData[grepl("summary of", stormData$EVTYPE, ignore.case = TRUE),], 
           n=5)

```

Upon this closer analysis, we can see that such instances have had the event type incorrectly recorded. The REMARKS column of the storm data contains a description of the event, which if needed, could be used to correctly populate the event type for such observations.

However, for the purposes of simple analysis, we will consider the EVTYPE values to be correct

#### Economic Effects

For the economic effects of severe weather, we must isolate what parts of the data define this.

Using the code book, ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)), we can see that the columns PROPDMG and CROPDMG indicate the value of damage to property and crops respectively. Each of those columns is followed by a corresponding *EXP column, which indicates an exponent (H-Hundreds, K-Thousands, M-Millions, B-Billions). 

In order to aggregate the property and crop damage costs per event type, a common unit will need to be used (in this case, whole dollars.) In order to do this, two additional columns *propertyDamage* and *cropDamage* will be introduced, which will be the *DMG value multiplied by the appropriate *DMGEXP value.

From the following analysis, we can determine that **Flood** is the severe weather event type that has the greatest effect on the cost of Property Damage and that **Drought** has the greatest effect on the cost of Crop Damage.

```{r EconomicEffects, echo=TRUE}

#subset data for crop damage cost, property damage cost (with exponents) and ev-type

stormDataEcon <- stormData[,c("EVTYPE", 
                              "PROPDMG", 
                              "PROPDMGEXP", 
                              "CROPDMG", 
                              "CROPDMGEXP")]

#create new vectors in the data frame
stormDataEcon$propertyDamage <- stormDataEcon$PROPDMG
stormDataEcon$cropDamage <- stormDataEcon$CROPDMG

#subset to rows only with damage
stormDataEcon <- stormDataEcon[stormDataEcon$PROPDMG != 0 | stormDataEcon$CROPDMG != 0,]

for (i in 1:nrow(stormDataEcon))
{
    if (stormDataEcon$PROPDMG[i] != 0)
    {
        if (stormDataEcon$PROPDMGEXP[i] == "H")
            stormDataEcon$propertyDamage[i] <- stormDataEcon$PROPDMG[i] * 10^2
        
        if (stormDataEcon$PROPDMGEXP[i] == "K")
            stormDataEcon$propertyDamage[i] <- stormDataEcon$PROPDMG[i] * 10^3
        
        if (stormDataEcon$PROPDMGEXP[i] == "M")
            stormDataEcon$propertyDamage[i] <- stormDataEcon$PROPDMG[i] * 10^6
        
        if (stormDataEcon$PROPDMGEXP[i] == "B")
            stormDataEcon$propertyDamage[i] <- stormDataEcon$PROPDMG[i] * 10^9    
    }
    
    if (stormDataEcon$CROPDMG[i] != 0)
    {
        if (stormDataEcon$CROPDMGEXP[i] == "H")
            stormDataEcon$cropDamage[i] <- stormDataEcon$CROPDMG[i] * 10^2
        
        if (stormDataEcon$CROPDMGEXP[i] == "K")
            stormDataEcon$cropDamage[i] <- stormDataEcon$CROPDMG[i] * 10^3
        
        if (stormDataEcon$CROPDMGEXP[i] == "M")
            stormDataEcon$cropDamage[i] <- stormDataEcon$CROPDMG[i] * 10^6
        
        if (stormDataEcon$CROPDMGEXP[i] == "B")
            stormDataEcon$cropDamage[i] <- stormDataEcon$CROPDMG[i] * 10^9
    }
}

#subset the relevant columns
stormDataEcon <- stormDataEcon[,c("EVTYPE",
                                  "propertyDamage",
                                  "cropDamage")]

stormDataEcon$EVTYPE <- factor(stormDataEcon$EVTYPE, 
                                   levels = stormDataEcon$EVTYPE)

#cost summation aggregrate by ev-type
economicImpactEventTypeAggregate <- aggregate(stormDataEcon,
                                              by = list(eventType = stormDataEcon$EVTYPE),
                                              FUN = function(x){sum(as.numeric(x))})

economicImpactEventTypeAggregate <- 
   economicImpactEventTypeAggregate[, c("eventType", 
                                        "propertyDamage",
                                        "cropDamage")]


#sort by totals
propertyDamageTop <- head(arrange(economicImpactEventTypeAggregate, 
                                             desc(propertyDamage)), 
                          n=15)

propertyDamageTop <- propertyDamageTop[,c("eventType", "propertyDamage")]
propertyDamageTop$eventType <- factor(propertyDamageTop$eventType, 
                                   levels = propertyDamageTop$eventType, 
                                   ordered = TRUE)

cropDamageTop <- head(arrange(economicImpactEventTypeAggregate, 
                                             desc(cropDamage)), 
                      n=15)

cropDamageTop <- cropDamageTop[,c("eventType", "cropDamage")]
cropDamageTop$eventType <- factor(cropDamageTop$eventType, 
                               levels = cropDamageTop$eventType, 
                               ordered = TRUE)

cropDamageTop
propertyDamageTop
```

```{r EconomicEffectsPlot, echo=TRUE, fig.align='center', fig.width=10, fig.height=5}

#plot dual plots

p1 <- qplot(eventType, 
            data = propertyDamageTop, 
            weight = propertyDamage, 
            binwidth = 1, 
            colour=I("blue"),
            fill=I("grey"),
            geom = "bar") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_y_continuous("Property Damage ($US)") + 
  xlab("Severe Weather Event Type") + 
  ggtitle("Property Damage Cost  vs\nSevere Weather Events Type\n(U.S. from 1995 - 2011)")

p2 <- qplot(eventType, 
            data = cropDamageTop, 
            weight = cropDamage, 
            binwidth = 1, 
            colour=I("blue"),
            fill=I("grey"),
            geom = "bar") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_y_continuous("Crop Damage ($US)") + 
  xlab("Severe Weather Event Type") + 
  ggtitle("Crop Damage Cost vs\nSevere Weather Events Type\n(U.S. from 1995 - 2011)")

grid.arrange(p1, p2, ncol = 2, main = "Economic Effects of Severe Weather Events")

```


#### Public Health Effects


For the health effects of severe weather, we must isolate what parts of the data define this.

Using the code book, ([Storm Events](http://ire.org/nicar/database-library/databases/storm-events/)), we can see that the columns INJURIES and FATALITIES indicate the number of injuries and fatalities caused by the event respectively. We will use these data characteristics to define the effect of severe weather events on public health.

From the following analysis, we can determine that **Tornado** is the severe weather event type that has the greatest effect on the number of injuries and on the number of fatalities.


```{r PublicHealthEffects, echo=TRUE}
#subset data for injuries, fatalities and ev-type
stormDataHealth <- stormData[,c("EVTYPE", "INJURIES", "FATALITIES")]

stormDataHealth$EVTYPE <- factor(stormDataHealth$EVTYPE, 
                                   levels = stormDataHealth$EVTYPE)

#summation aggregrate by ev-type
healthImpactEventTypeAggregate <- aggregate(stormDataHealth,
                                            by = list(eventType = stormDataHealth$EVTYPE),
                                            FUN = function(x){sum(as.numeric(x))})


healthImpactEventTypeAggregate <- 
   healthImpactEventTypeAggregate[, c("eventType", 
                                        "INJURIES",
                                        "FATALITIES")]

#sort by totals
injuriesTop <- head(arrange(healthImpactEventTypeAggregate, desc(INJURIES)), n=15)
injuriesTop <- injuriesTop[,c("eventType", "INJURIES")]
injuriesTop$eventType <- factor(injuriesTop$eventType, levels = injuriesTop$eventType, ordered = TRUE)

fatalitiesTop <- head(arrange(healthImpactEventTypeAggregate, desc(FATALITIES)), n=15)
fatalitiesTop <- fatalitiesTop[,c("eventType", "FATALITIES")]
fatalitiesTop$eventType <- factor(injuriesTop$eventType, levels = injuriesTop$eventType, ordered = TRUE)

injuriesTop
fatalitiesTop
```


```{r HealthEffectsPlot, echo=TRUE, fig.align='center', fig.width=10, fig.height=5}

#plot dual plots
p3 <- qplot(eventType, 
            data = injuriesTop, 
            weight = INJURIES, 
            binwidth = 1, 
            colour=I("blue"),
            fill=I("grey"),
            geom = "bar") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_y_continuous("Number of Injuries") + 
  xlab("Severe Weather Event Type") + 
  ggtitle("Number of Injuries  vs\nSevere Weather Events Type\n(U.S. from 1995 - 2011)")

p4 <- qplot(eventType, 
            data = fatalitiesTop, 
            weight = FATALITIES, 
            binwidth = 1, 
            colour=I("blue"),
            fill=I("grey"),
            geom = "bar") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  scale_y_continuous("Number of Fatalities") + 
  xlab("Severe Weather Event Type") + 
  ggtitle("Number of Fatalities  vs\nSevere Weather Events Type\n(U.S. from 1995 - 2011)")

grid.arrange(p3, p4, ncol = 2, main = "Health Effects of Severe Weather Events")

```

### Conclusion

Based on the above results, we can conclude that **tornado** is the severe weather most harmful with respect to population health, and that **flood** and **drought** have the greatest economic consequences.  

A future proposed improvement to this analysis, would be to rectify the issues described in the section **Event Type Analysis**, i.e. aggregate the data based on similar event types and clean up incorrectly recorded event types by getting the event type from the REMARKS column of such observations. 